El Mayor Postseismic \author{Trever Hines} \rule{\headwidth}{1.0pt}

### 2016-03-18 RoughDraft/ElMayor.tex
commit before push

Given the inability to describe both near and far field deformation with fault slip in an elastic lithosphere, \cite{Pollitz2012}, \cite{Rollins2015} and \cite{Spinler2015} have explored viscoelastic relaxation in the lower crust and upper mantle as a potential deformation mechanism. The lithospheric rheology is largely unknown and so modeling postseismic deformation with viscoelastic relaxation requires one to assume a rheologic model and find the best fitting model parameters, which is generally a computationally expensive nonlinear inverse problem. Consequently, a simplified structure for the lithosphere must be made to minimize the number of rheologic parameters that need to be estimated.  For example, it is commonly assumed that the lithosphere consits of only three homogeneously Maxwell viscoelastic layers, which may be an inadequate representation of the lithosphere \cite{Hines2013}\cite{Riva2009}. Additionally, it is necessary to make simplifying assumptions about the nature of afterslip.  For example \cite{Pollitz2012} and \cite{Spinler2015} assumed that afterslip does not persist beyond six months and one year respectively and subsequent displacements were assumed to be the result of only viscoelastic relaxation. However, numerous observations of interseismic fault creep has been recognized in this and similar tectonic settings and it has been speculated that such creep can be initiated as a postseismic process \cite{Cakir2012} \cite{Cetin2014}.
We are unaware of any oIn this study, we therefore do not make to implicit assumption that afterslip terminates after a given amount of time.  Indeed, the preferred viscoelastic model from \cite{Pollitz2012} underpredicts near field velocities, which could be indicative of unmodeled continued afterslip.

@@

We do not assume a parametric form for $u(t)$, (e.g. \cite{Rollins2015}), but rather we model $u(t)$ as integrated Brownian motion, so that 
\begin{equation}
    \dot{u}(t) = \sigma^2\int_0^t w(s) ds.
\end{equation}

@@

We show the difference in $u(t)$ for each station from 0 to 0.8 years (figure \ref{nearfield2} and \ref{farfield2}), 0.8 to 3.2 years (figure \ref{nearfield3} and \ref{farfield3}), and from 3.2 to 5.0 years (figure \ref{nearfield4} and \ref{farfield4}).  For far field stations (figure \ref{farfield2}, \ref{farfield3}, and \ref{farfield4}), we can see a clear south trending displacement throughout the first 3.2 years in stations as far as 400 km from of the El Mayor Cucapah epicenter.  These displacements are most pronounced along the direction of the El Mayor-Cucapah P and S axis, as would be expected for postseismic deformation.  Beyond 3.2 years, The southward trending far field postseismic deformation is barely perceptible. There is however, an eastward trend in far field stations just west of the Hector Mine epicenter.  We suspect this this deformation, which is inconsistent with the pattern expected for El Mayor-Postseismic, is the result of continued transient postseismic deformation following the Hector Mine earthquake.  The vertical deformation in the far field is difficult to attribute to postseismic processes.  Most far field stations display an initial subsidence for the first year after the El Mayor-Cucapah earthquake followed by continued uplift.  This trend in vertical deformation can be observed in all three of the qaudrants where postseismic data is available, which means that the vertical deformation does not exhibit an anti-symetric quadrant pattern, as would be expected for postseismic processes.  Although we use vertical deformation in our subsequent analysis, we do not put an emphasis on trying to describe the vertical deformation as it likely has non-tectonic origins.

@@

In this paper, we seek to find the mechanisms driving five years of postseismic deformation following the El Mayor-Cucapah. We consider afterslip and viscoelastic relaxation in the lithosphere and/or asthenosphere as candidate mechansism.  Poroelastic has also been used to model postseismic deformation (cite); however, the analysis of \cite{Gonzalez-ortega2014} suggests found that any contribution to postseismic deformation from poroelastic rebound would be negligable.  Additionally, we are considering stations which are sufficiently far away from the main rupture that poroelastic rebound should be insignificant.

In this study we estimate coseismic and time dependent postseismic fault slip, both which are assumed to occur on the fault geometry estimated by \cite{Wei2011a} with some modifications.  Field studies and other coseismic slip models have found a significantly more complicated geometry for the coseismically ruptured fault \cite{Fletcher2009} \cite{Oskin2012}; however, the relatively simple coseismic fault geometry from \cite{Wei2011a} is sufficient to describe the data in this study because most of the stations used here are sufficiently far from the epicenter that they are insensitive to the detailed fault geometry found by \cite{Fletcher2009} and \cite{Oskin2012}. We do, however, modify the fault geometry from \cite{Wei2011a} by extending segment 2 northward by 40 km which is motivated by the clustering of aftershocks on the northern tip of coseismic rupture zone \cite{Kroll2013} \cite{Hauksson2011}. This extended fault segment was also found to be necessary by \cite{Rollins2015} and \cite{Pollitz2012} for describing postseismic deformation.

To model postseismic deformation we must first assume a rheologic model.  The simplest viscoelastic rheologic model would be a Maxwell model, where shear stresses and strains can be described with the mechanical analogy of a spring and dashpot connected in series.  If coseismic and postseismic slip, described by $s(\xi,t)$, occurs on a fault $F$, and viscoelastic relaxation of subsequent stresses occurs in a Maxwell viscoelastic medium $L$, which has a spatially variable viscosity, $\eta(\zeta)$, then coseismic and postseismic deformation can be described by 
\begin{equation}\label{GeneralForward}
  u(x,t) = \int_F s(\xi,t)g(x,\xi)d\xi + 
           \int_0^t\int_F s(\tau,\xi) f(t-\tau,x,\xi) d\xi d\tau
\end{equation}
where $g(x,\xi)$ is the elastic Green's function relating slip at $\xi$ to instantaneous displacement at $x$ and $f(t,x,\xi)$ describes the rate of deformation at $x$ and time $t$ resulting from viscoelastic relaxation of stresses induced by slip  at $\xi$.  The variables which we are primarily concerned with estimating in this study are $s(\xi,t)$ and $\eta(\zeta)$.  Numerous experimental and geophysical studies suggest that a Maxwell rheology is inappropriate to describe the behaviour of the uppermost mantle \cite{Chopra1997} \cite{Pollitz2003} \cite{Freed2004} and so we do not limit ourselves to exploring Maxwell viscolesticity.  When considering other stress-linear viscoelastic models, such as a Zener or Burgers model, the description of displacements remains unchanged from eq \ref{GeneralForward} except that $f$ is parameterized by additional rheologic parameters which must also be estimated.  The difficulty in  determining an appropriate rheologic model and then estimating $s(\xi,t)$ and the unkown rheologic parameters from observations of $u(x,t)$ is that it is a nonlinear inverse problem with many unknown parameters and a computationally expensive forward problem.  We can greatly simplify the inverse problem by first considering the early postseismic period where we can approximate $u(x,t)$ for any stress-linear viscoelastic rheology as
\begin{equation}\label{ApproxForward}
  u(x,t) = \int_F s(\xi,t)g(x,\xi)d\xi + 
           \int_0^t\int_F\int_L \frac{s(\tau,\xi)}{\eta_\mathrm{eff}(\zeta)} h(x,\xi,\zeta) d\zeta d\xi d\tau
\end{equation}

@@

Numerous researchers have used vertical deformation   
as a discriminant for discerning whether afterslip or viscoelastic relaxation is the dominant postseismic mechanism (\cite{Hetland2014}, \cite{Pollitz2012} \cite{Rollins2015}).  As noted by \cite{Rollins2015} the uplift just north of the epicenter is suggestive of afterslip at seismogenic depths, as opposed to deeper afterslip which would predict subsidence north of the fault.

>>>>

Given the inability to describe both near and far field deformation with fault slip in an elastic lithosphere, \cite{Pollitz2012}, \cite{Rollins2015} and \cite{Spinler2015} have explored viscoelastic relaxation in the lower crust and upper mantle as a potential deformation mechanism. The lithospheric rheology is largely unknown and so modeling postseismic deformation with viscoelastic relaxation requires one to assume a rheologic model and find the best fitting model parameters, which is generally a computationally expensive nonlinear inverse problem. Consequently, a simplified structure for the lithosphere must be made to minimize the number of rheologic parameters that need to be estimated.  For example, it is commonly assumed that the lithosphere consists of only three homogeneously Maxwell viscoelastic layers, which may be an inadequate representation of the lithosphere \cite{Hines2013}\cite{Riva2009}. Additionally, it is necessary to make simplifying assumptions about the nature of afterslip. For example, one can assume a frictional model for afterslip and parameterize afterslip in terms its the unkown rheologic properties (e.g. \cite{Johnson2009} \cite{Johnson2004}). Additionally, one can assume that afterslip only dominates the first year of deformation (e.g. \cite{Pollitz2012} and \cite{Spinler2015}) and that viscoelastic relaxation is the dominant mechanism in later years. However, numerous observations of interseismic fault creep has been recognized in this and similar tectonic settings and it has been speculated that such creep can be initiated as a postseismic process \cite{Cakir2012} \cite{Cetin2014}.  We are unaware of any oIn this study, we therefore do not make to implicit assumption that afterslip terminates after a given amount of time.  Indeed, the preferred viscoelastic model from \cite{Pollitz2012} underpredicts near field velocities, which could be indicative of unmodeled continued afterslip.

@@

Studies of postseismic deformation typically assume a parametric form for $u(t)$ such as such as a logarithmic or exponential function. Such functions have been found empirically e.g. \cite{Savage2005a} and theoretically \cite{Marone1991} to adequately describe deformation resulting from postseismic deformation. However, by assuming a logarithmic or exponential form of $u(t)$ we run the risk of overfitting the GPS time series and inferring a nonexistent postseismic signal. We therefore do not assume any parameteric form for $u(t)$ and rather treat it as integrated Brownian motion, so that 
\begin{equation}
    \dot{u}(t) = \sigma^2\int_0^t w(s) ds.
\end{equation}

@@

We show the difference in $u(t)$ for each station from 0 to 0.8 years (figure \ref{nearfield2} and \ref{farfield2}), 0.8 to 3.2 years (figure \ref{nearfield3} and \ref{farfield3}), and from 3.2 to 5.0 years (figure \ref{nearfield4} and \ref{farfield4}).  For far field stations (figure \ref{farfield2}, \ref{farfield3}, and \ref{farfield4}), we can see a clear south trending displacement throughout the first 3.2 years in stations as far as 400 km from of the El Mayor Cucapah epicenter.  These displacements are most pronounced along the direction of the El Mayor-Cucapah P and S axis, as would be expected for postseismic deformation.  Beyond 3.2 years, The southward trending far field postseismic deformation is barely perceptible. There is however, an eastward trend in far field stations just west of the Hector Mine epicenter.  We suspect this this deformation, which is inconsistent with the pattern expected for El Mayor-Postseismic, is the result of continued transient postseismic deformation following the Hector Mine earthquake.  The vertical deformation in the far field is difficult to attribute to postseismic processes.  Most far field stations display an initial subsidence for the first year after the El Mayor-Cucapah earthquake followed by continued uplift.  This trend in vertical deformation can be observed in all three of the quadrants where postseismic data is available, which means that the vertical deformation does not exhibit an anti-symetric quadrant pattern, as would be expected for postseismic processes.  Although we use vertical deformation in our subsequent analysis, we do not put an emphasis on trying to describe the vertical deformation as it likely has non-tectonic origins.

@@

In this paper, we seek to find the mechanisms driving five years of postseismic deformation following the El Mayor-Cucapah earthquake. We consider afterslip and viscoelastic relaxation in the lithosphere and/or asthenosphere as candidate mechanisms.  Poroelastic rebound has also been used to model postseismic deformation (cite); however, \cite{Gonzalez-ortega2014} suggest that any contribution to postseismic deformation from poroelastic rebound would be negligable.  Additionally, we are considering stations which are sufficiently far away from the main rupture that poroelastic rebound should be insignificant.

In this study we estimate coseismic and time dependent postseismic fault slip, both of which are assumed to occur on the fault geometry estimated by \cite{Wei2011a} with some modifications.  Field studies \cite{Fletcher2014} and LIDAR observations \cite{Oskin2012} have revealed a significantly more complicated fault geometry than what was inferred by \cite{Wei2011a}, especially within the Sierra Cucapah.  However, we find the relatively simple coseismic fault geometry from \cite{Wei2011a} to be sufficient because most of the stations used in this station are sufficiently far from the El Mayor-Cucapah epicenter that they are insensitive to the details in the fault geometry found by \cite{Fletcher2014} and \cite{Oskin2012}.  We do, however, modify the fault geometry from \cite{Wei2011a} by extending segment 2 northward by 40 km which is motivated by the clustering of aftershocks on the northern tip of coseismic rupture zone \cite{Kroll2013} \cite{Hauksson2011}.  This extended fault segment was also found to be necessary by \cite{Rollins2015} and \cite{Pollitz2012} for describing postseismic deformation.

We consider a variety of rheologic models for the lithosphere and asthenosphere in this study.  The simplest rheologic model would be to consider the lithospere and asthenosphere to be effectively elastic. In such case, the rheologic parameters, consisting of the Lame parameters, $\lambda$ and $\mu$, and density, $\rho$, are reasonably well known, leaving the only unknowns to be the distribution of fault slip, which can be easily estimated due to the linearity of the forward problem. \cite{Rollins2015} found that postseismic deformation following the El Mayor-Cucapah earthquake can be well explained with afterslip on the coseismicly ruptured fault, which has been extended down to the lithosphere-asthenosphere boundary inferred by \cite{Lekic2011}, and they did not require any viscoelastic relaxation to describe the observations.  They did however, find that the amount of afterslip required for the southernmost fault segment to be unrealistically large, equivalent to a Mw7.2 earthquake.  Since most of the GPS data is located north of the rupture zone, slip on the Indiviso fault segment is least constrained and could be acting as a proxy for distributed relaxation in the lower crust or upper mantle.

It is well understood that afterslip at sufficiently great depths can explain postseismic deformation resulting from viscoelastic relaxation \cite{Savage1990}, and so in the interest of preventing a non-unique solution, we assuming afterslip to occur only at seismogenic depths (<15 km), and allow for viscoelastic relaxation in the lower crust (15-30 km) depth and the upper mantle.We choose such shallow constraints on afterslip because rate-state friction based models would suggest afterslip to occur at shallow depths to accommodate a coseismic slip deficit \cite{Marone1991}, which is indeed inferred in the coseismic models by \cite{Wei2011a} and (Fialko 2010). Additionally, as noted by \cite{Rollins2015} the postseismic uplift just north of the rupture zone can be well described with slip at seismogenic depths, while afterslip which is constrained to depths beneath 15 km predicts subsidence in that region.

A third reason for constraining slip to be shallower than 15 km depth is because we seek to estimate a lower crustal viscosity and trying to simultaneously estimate deep afterslip and a lower crustal viscosity leads creates a significant null space.  The reason for this is because viscoelastic relaxation within the lower crust of resulting from slip that is also within the lower crust will produce surface deformation in the direction opposite of slip. That is to say, the predicted viscoelastic deformation will cancel out the elastic deformation resulting in a trade off where a viscoelastic relaxation within a weak lower crust can be canceled out by slip at lower crustal depths.

With afterslip constrained to shallow depths, we must now select a viscoelastic rheology for the lower crust and upper mantle. Maxwell viscoelasticity, which can be schematically illustrated as a spring and dashpot connected in series, is the simplest and commonly used rheologic model for studies on postseismic deformation NUR IS SLS! (e.g. \cite{Nur1974}, \cite{Pollitz2000}, \cite{Hetland2003},\cite{Freed2006a}, \cite{Johnson2009}, \cite{Hearn2009}).  Assuming Maxwell viscoelasticity introduces a single the unknown rheologic parameter. A Maxwell viscoelastic model of the lithosphere would predict deformation that decays steadily over time, which fails to describe the multiple timescales of relaxation observed in postseismic deformation \cite{Savage1997}.  An additional deficiency in modeling the lithosphere with a Maxwell rheology is that inferred viscosities tend to depend on the the timescale of the geophysical process being modelled \cite{Peltier1981}. When assuming Maxwell viscoselasticity, upper mantle viscosities for Southern California inferred from postseismic deformation tend to be on the order of $10^{18}$ Pa s (\cite{Pollitz2000}, \cite{Thatcher2008}, \cite{Spinler2015}. However, studies of interseismic deformation \cite{Lundgren2009} and lake loading \cite{Luttrell2007}, which are both processes that occur on the timescale of $~10^2$ years, infer a Maxwell viscoelastic upper mantle with a viscosity on the order of $10^{19}$ Pa s.

To account for the apparently multiple timescales of relaxation, \cite{Pollitz2003} used a Burgers rheology upper mantle to model postseismic deformation following the Hector Mine earthquake and inferred a transient viscosity of $1.6\times10^{17}$ Pa s and a steady state viscosity of $4.6\times10^{18}$ Pa s, which still falls short of the steady state viscosity required for longer term geophysical processes. The particularly weak transient viscosity was found to be necessary to describe the early rapid postseismic deformation, although other researchers have successfully described the early postseismic deformation for the same earthquake with fault creep and poroelastic rebound. Nevertheless, laboratory experiments have demonstrated that mantle rock exhibits two phases of relaxation which can be adequately described with a Burgers rheology \cite{Chopra1997} and we should expect to see some postseismic signal resulting from transience viscoelasticity.

The third linear viscoelastic rheology which we consider is the Zener model, often referred to as a standard linear solid, whose mechanical analogue consists of a spring connected in series with a Kelvin element. A generalization of the Zener model, consisting of a series of Kelvin elements connected in series each of which representing a distinct deformation mechanism, is commonly used to describe the spectrum of frequency dependent attenuation \cite{Liu1976}.  The highest viscosity needed to describe attenuation of earth's lowest frequency normal modes is on the order of $10^{16}$ Pa s \cite{Yuen1982}, which is too weak to be perceptible in the GPS data used in the present study.

As an exploratory step, we also perform a slip inversion, assuming an elastic lithosphere and asthenosphere.  The difference between the analysis in  \cite{Rollins2015} and the present study is that we include a wider range of GPS stations in our inversion and we also perform a temporal estimation of afterslip, as opposed to estimating slip at a single epoch. The material properties used in our inversion are given in table 1 and are consistent with the properties used by \cite{Wei2011a} and the 1D SCEC background velocity model.  We discretize the

 and the constrained and the only para.  where the rheologic parameters       
In order to model postseismic deformation we must ddetermine an appropriate rheologic model for the lithosphere and asthenosphere.  
To model postseismic deformation we must first assume a rheologic model.  The simplest viscoelastic rheologic model would be a Maxwell model, where shear stresses and strains can be described with the mechanical analogy of a spring and dashpot connected in series.  If coseismic and postseismic slip, described by $s(\xi,t)$, occurs on a fault $F$, and viscoelastic relaxation of subsequent stresses occurs in a Maxwell viscoelastic medium $L$, which has a spatially variable viscosity, $\eta(\zeta)$, then coseismic and postseismic deformation can be described by 
\begin{equation}\label{GeneralForward}
  u(x,t) = \int_F s(\xi,t)g(x,\xi)d\xi + 
           \int_0^t\int_F s(\tau,\xi) f(t-\tau,x,\xi) d\xi d\tau
\end{equation}
where $g(x,\xi)$ is the elastic Green's function relating slip at $\xi$ to instantaneous displacement at $x$ and $f(t,x,\xi)$ describes the rate of deformation at $x$ and time $t$ resulting from viscoelastic relaxation of stresses induced by slip  at $\xi$.  The variables which we are primarily concerned with estimating in this study are $s(\xi,t)$ and $\eta(\zeta)$.  Numerous experimental and geophysical studies suggest that a Maxwell rheology is inappropriate to describe the behaviour of the uppermost mantle \cite{Chopra1997} \cite{Pollitz2003} \cite{Freed2004} and so we do not limit ourselves to exploring Maxwell viscolesticity.  When considering other stress-linear viscoelastic models, such as a Zener or Burgers model, the description of displacements remains unchanged from eq \ref{GeneralForward} except that $f$ is parameterized by additional rheologic parameters which must also be estimated.  The difficulty in  determining an appropriate rheologic model and then estimating $s(\xi,t)$ and the unkown rheologic parameters from observations of $u(x,t)$ is that it is a nonlinear inverse problem with many unknown parameters and a computationally expensive forward problem.  We can greatly simplify the inverse problem by first considering the early postseismic period where we can approximate $u(x,t)$ for any stress-linear viscoelastic rheology as
\begin{equation}\label{ApproxForward}
  u(x,t) = \int_F s(\xi,t)g(x,\xi)d\xi + 
           \int_0^t\int_F\int_L \frac{s(\tau,\xi)}{\eta_\mathrm{eff}(\zeta)} h(x,\xi,\zeta) d\zeta d\xi d\tau
\end{equation}

what this gives us is a prior constraint on lithospheric viscosity initial that is unbiased by neglecting the contribution of afterslip!

@@
